---
title: "Data types and other practice"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
---

Load packages
```{r load-packages}
library(tidyverse)
library(lubridate)
```

1. Using the air dataset, create a new variable called `date`. Assume the year is 2022. Make sure date is in a date format. Create a visualization showing Ozone over time.   
```{r air}
air <- as_tibble(airquality)
air %>%
  mutate(date = make_date(2022, Month, Day)) %>%
  ggplot(aes(x = date, y = Temp)) + geom_line()
```

2. Using the storms dataset, ask a question that can only be answered after doing some data wrangling (filtering, arranging, summarizing, selecting, mutating, etc.). Use the data to answer your question and create a visualization that supports your conclusion. Refer to the `storms` help file for more information about the dataset. 
```{r storms}
?storms

```

3. Using the midwest dataset, ask a question that can only be answered after doing some tidying (pivoting, spreading, uniting, etc.). Use the data to answer your question and create a visualization that supports your conclusion. Refer to the `midwest` help file for more information about the dataset. 
```{r midwest}
?midwest

```